Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-21 submission/manuscript.Rmd
discussion small edits

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity and identified substantial misexpression of testis, liver, muscle, and ovary-enriched genes within the _Kdm5c_-KO brain. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis. We demonstrated that while _Kdm5c_ mutant sex did not influence the whether sperm or egg genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. 
Germline-enriched mRNAs can become aberrantly expressed in _Kdm5c_-KO cells indirect of KDM5C, as KDM5C is enriched at germline gene promoters in EpiLCs but only bound to a subset of germline-enriched DEGs. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages. Furthermore, these results provide molecular footholds that can be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. KDM5C may therefore negatively regulate totipotency, since KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023]. However, _Kdm5c_-KO misexpression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023] and, unlike _Dazl_, KDM5C does not bind to the _Dux_ promoter. Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during differentiation into EpiLCs[@mochizukiRepressionGermlineGenes2021]. We found _Kdm5c_-KO EpiLCs expressed _Dazl_ but did not express 2-cell specific genes, suggesting the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression. Misexpression of germline genes in _Kdm5c_-KO EpiLCs could indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Altogether, these results demonstrate KDM5C repressess germline identity during embryogenesis and _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs.

Although eggs and sperm intially express very similar sets of germline genes for PGC formation and meiosis, some germline genes are sexually dimorphic. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of repression is independent of a given germline gene's sex. However, organismal sex did greatly influence the manifestation of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. X chromosome inactivation (XCI)  defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004] and loss of _Kdm5c_ impairs XCI[@samantaActivationXistEvolutionarily2022]. However, female germline DEGs were not significantly biased towards the X chromosome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.

It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promoters contained their consesus sequence. Additionally, some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Together, this suggests once germline genes like _Dazl_ and _Rfx2_ are activated in _Kdm5c_-KO cells, they can then turn on other ectopic germline programs downstream and loosely mimic germ cell development.

Unlike the previously characterized germline gene suppressors that place repressive histone modifications[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes the active histone mark histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression even more potently as their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. Additionally, we found _Kdm5c_-KOs accumulate H3K4me2/3 around the TSS of germline genes and impaired CpGme placement at CpG islands. KDM5C's shifting mode of germline gene repression as cells transition from na&#239;ve to primed pluripotency indicates chromatin regulators can employ both catalytic and non-catlytic repressive mechanisms at the same loci depending upon the developmental and cellular context.

- DNAme and CpG islands
	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
	- We found not all germline-enriched genes have CGIs, but many still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes
	- CGIs highly determine KDM5C recruitment to germline gene promoters.
			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function likely different from its somatic CGI function). 
	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
		- Other studies on germline gene repressors have shown they are importnat for CGIme, unclear if they participate in non-CGI TSS CpGme
	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that can impair learning and memory when overexpressed in the mouse brain[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Thus, KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_. Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. Ectopic germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and can drive brain turmor formation in _Drosophila_[@janicEctopicExpressionGermline2010], indicating they may promote genome instability and cellular de-differentiation. Intriguingly, select models of other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation However, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.

>>>>

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
We first characterized the misexpression of tissue-enriched genes within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testis somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis. We demonstrated that while _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. 
Germline-enriched mRNAs can become aberrantly expressed in _Kdm5c_-KO cells indirect of KDM5C, as although KDM5C is enriched at germline gene promoters in EpiLCs, it is only bound to a subset of germline-enriched DEGs. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages. Furthermore, these data provide molecular footholds that can then be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. KDM5C may therefore negatively regulate totipotency, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023]. However, misexpression of _Dazl_ in _Kdm5c_-KO ESCs was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023] and, unlike _Dazl_, KDM5C does not bind to the _Dux_ promoter.

Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs also expressed _Dazl_, they did not express 2-cell specific genes. Misexpression of germline genes in _Kdm5c_-KO EpiLCs could indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.

Although eggs and sperm employ the same transcriptional programs for shared functions, e.g. PGC formation,  meiosis, and genome defense, some germline genes are sex specific. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex. However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. X chromosome inactivation (XCI) defects could explain why _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004] and loss of _Kdm5c_ impairs XCI[@samantaActivationXistEvolutionarily2022]. However, female germline DEGs were not significantly biased towards the X chromosome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.

It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007]. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence. Additionally, some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.

Unlike the previously characterized germline gene suppressors that place repressive histone modifications[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes the active histone mark histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at germline CpG islands. KDM5C's shifting role in germline gene repression during the transition from na&#239;ve to primed pluripotency indicates chromatin modifiers can switch between catalytic and non-catlytic gene regulatory mechanisms at the same loci over development.

- DNAme and CpG islands
	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
	- We found only 30% of germline-enriched genes have CGIs, but the majority of CGI-free germline genes still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes.
	- CGIs highly determine KDM5C recruitment to germline gene promoters.
			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms

The above work provides the mechanistic foundation for KDM5C's repression of germline-specific transcription and offers novel insight into how dysregulation of _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022]. _APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008]. KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_. Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. Ectopic germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and can drive brain turmor formation in _Drosophila_[@janicEctopicExpressionGermline2010], indicating their dysregulation may promote genome instability and cellular de-differentiation. Intriguingly, some models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation However, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
